---
title: "Pipeline Accident Analysis"
author: "Jiayang (Lyra) Wang"
date: "7/17/2022"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

```{r packages, include=FALSE}
# packages
library(dplyr)
library(ggplot2)
library(readxl)
library(tidyr)
library(extrafont)
library(tidyverse)
library(poorman)
library(sf)
library(ggmap)
library(rlang)
library(rjson)
library(magrittr)
library(mapview)
library(scales)
library(usmap)
library(maps)
library(mapdata)
library(stringr)
library(ggforce)
library(giscoR)

install.packages("giscoR")

# import font
loadfonts(quiet = T)
windowsFonts("Franklin Gothic Book" = windowsFont("Franklin Gothic Book"))

# Bootstrapping Function Definition
sample_mean <- function(data, indices){
  sample <- data[indices]
  bar <- mean(sample)
  return(bar)
}
```

```{r data import}
# data import - "Oil Pipeline Accidents, 2010 - Present" data set from Kaggle
df <- read.csv("C:/Users/wangly0/Desktop/database.csv")
```

Pipeline incidents lead to severe social and environmental costs. In this document, we aim to analyze the cause for crude oil pipeline incidents and discuss policy progress and potential improvements. 

The first step of this analysis is to understand the data set on hand.  1) what are the variables; 2) missing value issues
```{r column name, echo=FALSE, message=TRUE}
# display column names of this data set
colnames(df)

# missing data analysis
dfNA <- df %>% 
  # identify missing values with NA - not all missing values were marked with NA
  mutate(across(everything(), ~replace(., . == "", NA))) %>% 
  # count the number of missing values in the data set
  summarise_all(~sum(is.na(.))) %>% 
  # create data frame with two columns: variable in the data set, count of NA values in the variable
  gather(Variable, NACount, 1:48, factor_key = TRUE) %>% 
  # remove variables that do not have NA values
  filter(NACount != 0) %>% 
  # arrange variables by descending order of NA values
  arrange(-NACount) %>% 
  # calculate percentage of missing value 
  mutate(Percentage = round(NACount/nrow(df),3))

dfNA

write.csv(dfNA, "C:/Users/wangly0/Desktop/NA.csv",row.names = FALSE)
  
# replace NA values with zero and binary (yes/no) values
df <- df %>% 
  # replace NA values in fatalities, injuries, release barrels, and costs with 0
  mutate_at(vars(Operator.Employee.Fatalities,
                 Operator.Contractor.Fatalities,
                 Emergency.Responder.Fatalities,
                 Other.Fatalities,
                 Public.Fatalities,
                 All.Fatalities,
                 Operator.Employee.Injuries,
                 Operator.Contractor.Injuries,
                 Emergency.Responder.Injuries,
                 Other.Injuries,
                 Public.Injuries,
                 All.Injuries,
                 Intentional.Release..Barrels.,
                 Other.Costs,
                 Public.Private.Property.Damage.Costs,
                 Environmental.Remediation.Costs,
                 Property.Damage.Costs,
                 Emergency.Response.Costs,
                 Lost.Commodity.Costs), 
            ~replace_na(., 0)) %>% 
  # replace NA values as binary "NO" for variable
  mutate_at(vars(Pipeline.Shutdown),
            ~replace_na(.,"NO"))
```
There are a few observations under offshore that are outside of US boundaries. Figure below shows the 
```{r}
# replace NA values in Accident.State based on coordinate information 
dfGeo <- df %>% 
  filter(Pipeline.Location == "OFFSHORE") %>% 
  filter(Accident.State == "") %>% 
  select(Accident.Latitude, Accident.Longitude)

dfGeo1 <- df %>% 
  filter(Pipeline.Location == "OFFSHORE") %>% 
  filter(Accident.State != "") %>% 
  select(Accident.Latitude, Accident.Longitude)

mapview(dfGeo, xcol =  "Accident.Longitude", ycol = "Accident.Latitude", crs = 4269, grid = FALSE, col.regions = "red", layer.name = c("NA Value"))+
  mapview(dfGeo1, xcol =  "Accident.Longitude", ycol = "Accident.Latitude", crs = 4269, grid = FALSE, col.regions = "blue", layer.name = c("Louisiana"))

# replace NA with "Gulf of Mexico"
df<- df %>% 
  mutate(Accident.State = ifelse(Accident.State == "", "Gulf of Mexico", Accident.State))

# create a map view of all the crude oil incidents in the data set
  mapview(df[df$Liquid.Type == "CRUDE OIL" & df$Pipeline.Location == "ONSHORE",], xcol =  "Accident.Longitude", ycol = "Accident.Latitude", crs = 4269, grid = FALSE, col.regions = "#ffbf00", layer.name = c("Onshore")) +
  mapview(df[df$Liquid.Type == "CRUDE OIL" & df$Pipeline.Location == "OFFSHORE",],xcol =  "Accident.Longitude", ycol = "Accident.Latitude", crs = 4269, grid = FALSE, col.regions = "darkgreen", layer.name = c("Offshore"))
  
# remove one wrong coordinates from the data set and create crude oil only data set
dfCO <- df %>% 
    # remove coordinates that land in Asia
    filter(Accident.Latitude != 41.890476) %>% 
    # remove coordinates that land in Canada
    filter(Accident.Latitude != 50.839319) %>% 
    filter(Liquid.Type == "CRUDE OIL")

# map view of crude oil incidents after removing ones that do not belong to the US
mapview(dfCO[dfCO$Pipeline.Location == "ONSHORE",], xcol =  "Accident.Longitude", ycol = "Accident.Latitude", crs = 4269, grid = FALSE, col.regions = "#ffbf00", layer.name = c("Onshore")) +
  mapview(dfCO[dfCO$Pipeline.Location == "OFFSHORE",],xcol =  "Accident.Longitude", ycol = "Accident.Latitude", crs = 4269, grid = FALSE, col.regions = "darkgreen", layer.name = c("Offshore"))

# calculate the percentage of onshore and offshore crude oil incidents 
dfCO %>% 
  group_by(Pipeline.Location) %>% 
  tally() %>% 
  mutate(Percentage = round(n/sum(n),2))


```
Severity of crude oil pipeline accidents - total cost
```{r}
# check out the distribution of all costs
dfCO %>% 
  select(All.Costs) %>% 
  summary()

# calculate the total cost of crude oil damage
dfCO %>% 
  summarise(Total.Cost = sum(All.Costs))

# total cost of crude oil spills 
dfCO_Cost <- dfCO %>% 
  filter(Accident.State != "Gulf of Mexico") %>% 
  group_by(Accident.State) %>% 
  summarise(Total.Cost = sum(All.Costs))

test <- dfCO_Cost %>% 
  rename(abbr = Accident.State) %>% 
  right_join(statepop, by = "abbr") %>% 
  select(fips, abbr, full, Total.Cost)

plot_usmap(data = test, values = "Total.Cost", color = "red") + 
  scale_fill_continuous(
    low = "white", high = "darkblue", name = "Total Cost", label = scales::comma
  ) + theme(legend.position = "right")


```
Severity of crude oil pipeline accident at state level
```{r}
# Classify states by cost
dfCO_State <- dfCO %>% 
  group_by(Accident.State) %>% 
  summarise(Accident.Count = n(),
            Property.Damage.Costs = sum(Property.Damage.Costs),
            Lost.Commodity.Costs = sum(Lost.Commodity.Costs),
            Public.Private.Property.Damage.Costs = sum(Public.Private.Property.Damage.Costs),
            Emergency.Response.Costs = sum(Emergency.Response.Costs),
            Environmental.Remediation.Costs =sum(Environmental.Remediation.Costs),
            Other.Costs = sum(Other.Costs),
            Total.Cost = sum(All.Costs),
            Average.Cost = Total.Cost/Accident.Count) %>% 
  arrange(-Total.Cost) %>% 
  mutate(Size = ifelse(Total.Cost >100000000, ">100M",
                       ifelse(Total.Cost <100000000 & Total.Cost>10000000, "10M ~ 100M",
                              ifelse(Total.Cost<10000000 & Total.Cost>1000000, "1M ~ 10M", "<1M"))))
dfCO_State
dfCO_State %>% 
  ungroup() %>% 
  select(-Total.Cost) %>% 
  gather(Cost.Category, Cost, Property.Damage.Costs:Average.Cost, factor_key = TRUE) %>% 
  mutate(Size = factor(Size, levels = c(">100M", "10M ~ 100M", "1M ~ 10M", "<1M"))) %>% 
  ggplot()+
  geom_bar(mapping = aes(x = reorder(Accident.State, -Cost), y = Cost, fill = Cost.Category),position = "stack", stat = "identity", width = 0.9)+
  facet_row(vars(Size), scales = "free", space = "free")+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 8))+
  scale_fill_brewer(palette = "Set3")+
  labs(x = "", y = "Costs")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "bottom",
        legend.title = element_blank(),
        text = element_text(size = 15, family = "Franklin Gothic Book"))

ggsave(filename = "Cost.png", path = "C:/Users/wangly0/Desktop", height = 6, width = 16)
```

Severity by looking at how common such things happen
```{r}
dfCO_State %>% 
  ggplot()+
  geom_bar(mapping = aes(x = reorder(Accident.State, -Average.Cost), y = Average.Cost, fill = Accident.State), stat = "identity")+
  geom_text(mapping = aes(x = reorder(Accident.State, -Average.Cost), y = Average.Cost,label = Accident.Count), vjust = -0.2, size = 6)+
  coord_cartesian(ylim = c(0, 17000000))+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 8))+
  labs(x = "", y="Average Cost per Accident")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "none",
        legend.title = element_blank(),
        text = element_text(size = 18, family = "Franklin Gothic Book"))

ggsave(filename = "CostPerAccident.png", path = "C:/Users/wangly0/Desktop", height = 6, width = 16)

dfCO_State %>% 
  arrange(-Average.Cost)
```

Distribution of accidents
```{r}
dfCO_State$Total.Cost <- as.numeric(dfCO_State$Total.Cost)

dfCO

dfCO %>% 
  mutate(Pipeline.Type = ifelse(Pipeline.Type == "" & Pipeline.Location == "OFFSHORE", "OFFSHORE", Pipeline.Type)) %>% 
  arrange(-All.Costs) %>% 
  ungroup() %>% 
  mutate(CumSum = cumsum(All.Costs)/sum(All.Costs),
         CumCount = seq.int(nrow(dfCO))/nrow(dfCO)) %>%  
  ggplot()+
  geom_point(mapping = aes(x = CumCount, y = CumSum, color = Pipeline.Type), size = 3)+
  scale_color_brewer(palette = "Dark2")+
  labs(x = "Cumulative Count of Accidents", y = "Cumulative Cost of Accidents")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "top",
        legend.title = element_blank(),
        text = element_text(size = 18, family = "Franklin Gothic Book"))

ggsave(filename = "CostDistributionPipelineType.png", path = "C:/Users/wangly0/Desktop", height = 5, width = 9)
```

Histogram of accident by pipeline type
```{r}
dfCO %>% 
  mutate(Pipeline.Type = ifelse(Pipeline.Type == "" & Pipeline.Location == "OFFSHORE", "OFFSHORE", Pipeline.Type)) %>% 
  arrange(-All.Costs) %>% 
  ungroup() %>% 
  select(Pipeline.Type, All.Costs) %>% 
  mutate(Count = 1) %>% 
  group_by(Pipeline.Type) %>% 
  summarise(Total.Cost = sum(All.Costs), Total.Count = sum(Count)) %>% 
  gather(Category, Value, Total.Cost:Total.Count) %>% 
  ggplot()+
  geom_bar(mapping = aes(x = Category,y = Value, fill = Pipeline.Type), position = "fill", stat = "identity")+
  scale_fill_brewer(palette = "Dark2")+
  labs(x = "", y = "")+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none",
        text = element_text(size = 18, family = "Franklin Gothic Book"))

ggsave(filename = "CostCountPipelineType.png", path = "C:/Users/wangly0/Desktop", height = 3, width = 6)

# percentage of each pipeline type
dfCO %>% 
  mutate(Pipeline.Type = ifelse(Pipeline.Type == "" & Pipeline.Location == "OFFSHORE", "OFFSHORE", Pipeline.Type)) %>% 
  arrange(-All.Costs) %>% 
  ungroup() %>% 
  select(Pipeline.Type, All.Costs) %>% 
  mutate(Count = 1) %>% 
  group_by(Pipeline.Type) %>%
  summarise(Total.Cost = sum(All.Costs), 
            Total.Count = sum(Count)) %>% 
  ungroup() %>% 
  mutate(Total.Cost.Per = Total.Cost/sum(Total.Cost),
         Total.Cound.Per = Total.Count/sum(Total.Count),
         Cost.Accident = Total.Cost/Total.Count)
```

Analysis on cause of crude oil pipeline accidents
```{r}
dfCO %>% 
  select(Pipeline.Location,Pipeline.Type, Cause.Category:Net.Loss..Barrels., Property.Damage.Costs:All.Costs) %>% 
  group_by(Cause.Category, Cause.Subcategory) %>% 
  summarise(Total.Cost = sum(All.Costs), 
            Count = n(),
            Cost.Accident = Total.Cost/Count) %>% 
  arrange(-Total.Cost)

mapview(dfCO, xcol =  "Accident.Longitude", ycol = "Accident.Latitude",zcol = "Cause.Category", size = "All.Costs", crs = 4269, grid = FALSE, col.regions = c("darkgreen", "yellow", "darkblue", "orange", "gray","purple","lightblue"))

dfCO %>% 
  group_by(Cause.Category) %>% 
  summarise(Total.Cost = sum(All.Costs), 
            Count = n()) %>% 
  mutate_at(vars("Total.Cost", "Count"), as.numeric) %>% 
  ungroup() %>% 
  mutate(Cost.Accident = Total.Cost/Count,
         Total.Cost.Per = Total.Cost/sum(Total.Cost)) %>% 
  ggplot()+
  geom_bar(mapping = aes(x = "", y = Total.Cost.Per, fill = Cause.Category), stat = "identity", width = 1, color = "white")+
  coord_polar("y", start = 0)+
  labs(x = "", y = "")+
  scale_fill_manual(values = c("darkgreen", "yellow", "darkblue", "orange", "gray","purple","lightblue"))+
  theme_void()+
  theme(legend.position = "none",
        text = element_text(size = 18, family = "Franklin Gothic Book"))

ggsave(filename = "MainCausePie.png", path = "C:/Users/wangly0/Desktop", height = 6, width = 6)

dfCO %>% 
  group_by(Cause.Category) %>% 
  summarise(Total.Cost = sum(All.Costs), 
            Count = n()) %>% 
  mutate_at(vars("Total.Cost", "Count"), as.numeric) %>% 
  ungroup() %>% 
  mutate(Cost.Accident = Total.Cost/Count,
         Total.Cost.Per = Total.Cost/sum(Total.Cost)) %>% 
  arrange(-Total.Cost)
```

Material/weld/equip failure analysis
```{r}
dfCO %>% 
  group_by(Cause.Category,Cause.Subcategory) %>% 
  summarise(Total.Cost = sum(All.Costs), 
            Count = n()) %>% 
  mutate_at(vars("Total.Cost", "Count"), as.numeric) %>% 
  ungroup() %>% 
  mutate(Cost.Accident = Total.Cost/Count,
         Total.Cost.Per = Total.Cost/sum(Total.Cost)) %>% 
  arrange(-Cost.Accident)

dfCO %>% 
  filter(Cause.Category == "MATERIAL/WELD/EQUIP FAILURE") %>% 
  group_by(Cause.Subcategory) %>% 
  summarise(Total.Cost = sum(All.Costs), 
            Count = n()) %>% 
  mutate_at(vars("Total.Cost", "Count"), as.numeric) %>% 
  ungroup() %>% 
  mutate(Cost.Accident = Total.Cost/Count,
         Total.Cost.Per = Total.Cost/sum(Total.Cost)) %>% 
  arrange(-Cost.Accident) %>% 
  ggplot()+
  geom_bar(mapping = aes(x = "", y = Total.Cost.Per, fill = Cause.Subcategory), stat = "identity", width = 1, color = "white")+
  coord_polar("y", start = 0)+
  labs(x = "", y = "")+
  theme_void()+
  theme(legend.position = "none",
        text = element_text(size = 18, family = "Franklin Gothic Book"))

ggsave(filename = "EquipFailurePie.png", path = "C:/Users/wangly0/Desktop", height = 6, width = 6)

dfCO %>% 
  filter(Cause.Category == "MATERIAL/WELD/EQUIP FAILURE") %>% 
  group_by(Cause.Subcategory) %>% 
  summarise(Total.Cost = sum(All.Costs), 
            Count = n()) %>% 
  mutate_at(vars("Total.Cost", "Count"), as.numeric) %>% 
  ungroup() %>% 
  mutate(Cost.Accident = Total.Cost/Count,
         Total.Cost.Per = Total.Cost/sum(Total.Cost),
         Total.Count.Per = Count/sum(Count)) %>% 
  arrange(-Cost.Accident)
```
improvement observed during the past 7 years
```{r}
dfCO %>% 
  filter(Accident.Year != 2017) %>% 
  group_by(Accident.Year) %>% 
  summarise(Total.Cost = sum(All.Costs),
            Count = n(),
            Average.Cost = Total.Cost/Count) %>% 
  ungroup() %>% 
  mutate_at(vars(Total.Cost), as.numeric) %>% 
  mutate(Total.Cost.Scale = Total.Cost/Total.Cost[Accident.Year == 2010],
         Count.Scale = Count/Count[Accident.Year == 2010],
         Average.Cost.Scale = Average.Cost/Average.Cost[Accident.Year == 2010]) %>% 
  ggplot()+
  geom_line(mapping = aes(x = Accident.Year, y = Total.Cost.Scale, color = "Total.Cost"), size = 1) +
  geom_line(mapping = aes(x = Accident.Year, y = Average.Cost.Scale, color = "Average.Cost"), size = 1)+
  geom_line(mapping = aes(x = Accident.Year, y = Count.Scale, color = "Count"), size = 1)+
  scale_color_manual(name = "", values = c("Total.Cost" = "#ffbf00", "Average.Cost" = "darkgreen", "Count" = "#bf0000"))+
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017))+
  labs(x = "", y = "Changes Compared to 2010 Baseline")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = c(0.15,0.85),
        legend.title = element_blank(),
        legend.background = element_blank(),
        legend.box.background = element_rect(color = "black"),
        text = element_text(size = 16, family = "Franklin Gothic Book"))

ggsave(filename = "Improvement.png", path = "C:/Users/wangly0/Desktop", height = 6, width = 8)
```

Material/weld/equipment failure change over the years
```{r}
dfCO %>% 
  filter(Cause.Category == "MATERIAL/WELD/EQUIP FAILURE") %>% 
  group_by(Accident.Year, Cause.Subcategory) %>% 
  summarise(Total.Cost = sum(All.Costs),
            Count = n(),
            Average.Cost = Total.Cost/Count) %>% 
  ungroup() %>% 
  mutate_at(vars(Total.Cost, Count), as.numeric) %>% 
  ggplot()+
  geom_bar(mapping = aes(x = Accident.Year, y = Total.Cost, fill = Cause.Subcategory), stat = "identity", position = "stack")+
  labs(x = "", y = "Total Cost")+
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017))+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = c(0.6,0.6),
        legend.title = element_blank(),
        legend.background = element_blank(),
        legend.box.background = element_rect(color = "black"),
        text = element_text(size = 16, family = "Franklin Gothic Book"))

ggsave(filename = "ImprovementCause.png", path = "C:/Users/wangly0/Desktop", height = 6, width = 8)

dfCO %>% 
  filter(Cause.Category == "MATERIAL/WELD/EQUIP FAILURE") %>% 
  arrange(-All.Costs)

dfCO %>% 
  arrange(-All.Costs) %>% 
  slice(1:10) %>% 
  group_by(Accident.Year) %>% 
  tally()
```
barrels lost compared to barrels transported
```{r}
Trans <- data.frame(Accident.Year = c(2010,2011,2012,2013,2014,2015,2016,2017),
                    Barrels = c(593535000, 518308000, 561100000,599580000, 707823000, 918703000, 1000827000, 1183714000))

# dfCO %>% 
#   filter(Accident.Year != 2017) %>% 
#   group_by(Accident.Year) %>% 
#   summarise(Release = sum(Unintentional.Release..Barrels.,Intentional.Release..Barrels.),
#             Recovery = sum(Liquid.Recovery..Barrels.),
#             Net.Loss = sum(Net.Loss..Barrels.)) %>% 
#   ungroup() %>% 
#   left_join(Trans, by = "Accident.Year") %>% 
#   mutate(Lost.Per = Net.Loss/Barrels) %>% 
#   ggplot()+
#   geom_line(mapping = aes(x = Accident.Year, y = Barrels, color = "Transported (Barrels)"))+
#   geom_line(mapping = aes(x = Accident.Year, y = Release, color = "Released (Barrels)"))+
#   geom_line(mapping = aes(x = Accident.Year, y = Recovery, color = "Recovered (Barrels)"))+
#   geom_line(mapping = aes(x = Accident.Year, y = Net.Loss, color = "Net Loss (Barrels)"))+
#   # scale_y_continuous(label = scales::percent)+
#   scale_color_manual(name = "", values = c("Transported (Barrels)" = "#ffbf00",
#                                            "Released (Barrels)" = "purple",
#                                            "Recovered (Barrels)" = "red",
#                                            "Net Loss (Barrels)" = "darkgreen"))


dfCO %>% 
  filter(Accident.Year != 2017) %>% 
  group_by(Accident.Year) %>% 
  summarise(Net.Loss = sum(Net.Loss..Barrels.)) %>% 
  ungroup() %>% 
  left_join(Trans, by = "Accident.Year") %>% 
  mutate(Lost.Per = Net.Loss/Barrels) %>% 
  ggplot()+
  geom_line(mapping = aes(x = Accident.Year, y = Lost.Per), size = 1)+
  scale_y_continuous(label = scales::percent)+
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017))+
  labs(x = "", y = "Percent Transported Lost to Accidents")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.title = element_blank(),
        text = element_text(size = 16, family = "Franklin Gothic Book"))

ggsave(filename = "Percent.png", path = "C:/Users/wangly0/Desktop", height = 6, width = 8)
```





